Parsing Postfix log files

### 2007-11-06 doc/logparser.tex
Added a glossary and acronym list. Started working on Carl's corrections.

% Useful stuff for math mode.
\usepackage{amstext}
% Wrapping of URLs.  This is used in the bibliography instead of \url{}.
\usepackage{url}
%\DeclareUrlCommand\urlwrap{}
%\newcommand{\urlwrapfill}[1]{\urlwrap{#1}\hfill}
\newcommand{\parsername}{ASO}
% Include images
\usepackage[final]{graphicx}
% Change how nested enumerate environments are labelled.
\renewcommand{\labelenumii}{\roman{enumii}:}
\renewcommand{\refname}{Bibliography}
% Add the bibliography into the table of contents.
\usepackage[section,numbib]{tocbibind}
% When creating a PDF make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.  The title and
% author don't work - I think either gnuplot or graphviz clobbers it.
\usepackage{hyperref}
\hypersetup{
    pdftitle    = {Parsing Postfix log files},
    pdfauthor   = {John Tobin},
    final       = true,
    pdfborder   = {0 0 0},
%    breaklinks  = true,
}
\usepackage{breakurl}
\usepackage{lastpage}

@@

% Replacement for \ref{}, adds the page number too.
\newcommand{\refwithpage}[1]{%
    \empty{}\ref{#1} (page~\pageref{#1})%
}

@@

\title{Parsing Postfix log files}
\author{John Tobin \\ School of Computer Science and Statistics \\ 
Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
\date{}
\maketitle

@@

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or check whether their
new anti-spam measures are working.  The more adventurous will have
generated statistics to show how many hits each of their anti-spam measures
has gotten in the last week, and possibly even generated some graphs to
clearly illustrate the point to management or users.\footnote{This was the
author's first real foray into processing Postfix logs.}  Very few will
have performed in-depth parsing and analysis of their logs, where the
parsing must correlate the log lines per-connection or per-queueid rather
than processing lines independently.  One of the barriers to this type of
processing is the unstructured nature of Postfix logs, where each log line
was added on an ad hoc basis as a requirement was discovered or new
functionality was added.  Further complication arises because the set of
rejection messages is not fixed: new messages can be added by the
administrator with custom checks; every Real-time Black List returns a
different explanatory message; policy servers may log different messages
depending on the characteristics of the connection; there are many ways in
which the log lines may differ between servers.  This paper documents the
difficult process of parsing Postfix logs, presenting a program which
parses logs and places the resulting data into a database.  The gathered
data can then be used to optimise current anti-spam measures, provide a
baseline to test new anti-spam measures against, or to produce statistics
showing how effective those measures are.  There are numerous other uses
for such data: improving server performance by identifying troublesome
destinations and reconfiguring appropriately or dedicating transports to
those destinations; identifying regular high volume uses (e.g.\ customer
newsletters) and restricting those uses to off-peak times; detecting virus
outbreaks which propagate via email; as a base for billing customers on a
shared server.  Preserving the raw data enables users to develop a
multitude of uses far beyond those conceived of by the author.

@@

Section~\ref{rules} discusses the parsing rules in detail, describing the
table used to hold the rules and explaining their structure, using an
example rule and sample data it matches successfully against, followed by a
discussion of rule efficiency concerns.

@@

Section~\ref{other-parsers} discusses other parsers and why they were
deemed unsuitable for the task.

@@

Section~\ref{graphs} contains graphs illustrating the topics discussed in
section~\ref{rule efficiency}, rule efficiency.

@@

This section provides background information helpful in understanding the
remainder of the paper.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an API\@; a brief introduction to the SMTP protocol; and a
longer introduction to Postfix, concentrating on the topics most relevant
to this paper, namely Postfix restrictions and policy servers.  The
assumptions made in designing and implementing the parser are explained, as
are the conventions used in this paper.  Other projects which attempt to
parse Postfix logs are summarised (full details are available in
section~\refwithpage{other-parsers}), plus a review of previously published
research in this area.

@@

@@

\subsection{SMTP background}
\label{smtp background}

The Simple Mail Transfer Protocol, originally defined in RFC
821~\cite{RFC821} and updated in RFC 2821~\cite{RFC2821}, is used for
transferring mail between Mail Transport Agents (MTAs).  It is a simple,
human readable, plain text protocol, making it quite easy to test and debug
problems with it.  Despite the simplicity many virus and/or spam sending
programs fail to implement it properly, so requiring strict adherence to
the protocol specification is beneficial in protecting against spam and
viruses.\footnote{This violates the principle of \textit{Be liberal in what
you accept, and conservative in what you send\/} from RFC
760~\cite{rfc760}, but unfortunately that principle was written in a
friendlier time.}  A typical SMTP conversation resembles the following (the
lines starting with a three digit number are sent by the server, all other
lines are sent by the client):

@@

            \item Modify \texttt{smtpd\_helo\_restrictions} in
                \texttt{/etc/postfix/main.cf} to contain\newline
                \hspace*{2em}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}

@@

        The purpose of logging the HELO in
        \texttt{smtpd\_helo\_restrictions} is to log only once per mail; it
        also seems like the natural place.

@@

>>>>

% Useful stuff for math mode.
\usepackage{amstext}
% Wrapping of URLs.  This is used in the bibliography instead of \url{}.
\usepackage{url}
\newcommand{\parsername}{ASO}
% Include images
\usepackage[final]{graphicx}
% Change how nested enumerate environments are labelled.
\renewcommand{\labelenumii}{\roman{enumii}:}
\renewcommand{\refname}{Bibliography}
% Add the bibliography into the table of contents.
\usepackage[section,numbib]{tocbibind}

\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
% Redefine the command used to produce the glossary title, because the
% default command produces an unnumbered section whereas I want a numbered
% section.
\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
\makeglossary{}
\makeacronym{}

% When creating a PDF make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.  The title and
% author don't work - I think either gnuplot or graphviz clobbers it.
\usepackage{hyperref}
\hypersetup{
    pdftitle    = {Parsing Postfix log files},
    pdfauthor   = {John Tobin},
    final       = true,
    pdfborder   = {0 0 0},
%    breaklinks  = true,
}
\usepackage{breakurl}
\usepackage{lastpage}

@@

% Replacement for \ref{}, adds the page number too.
\newcommand{\refwithpage}[1]{%
    \empty{}\ref{#1} [page~\pageref{#1}]%
}

@@

% Pull in the acronyms early, so they can be used throughout the text.
\input{logparser-acronyms.tex}

\title{Parsing Postfix log files}
\author{John Tobin \\ School of Computer Science and Statistics \\ 
Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
\maketitle

@@

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or check whether their
new anti-spam measures are working.  The more adventurous will have
generated statistics to show how many hits each of their anti-spam measures
has gotten in the last week, and possibly even generated some graphs to
clearly illustrate the point to management or users.\footnote{This was the
author's first real foray into processing Postfix logs.}  Very few will
have performed in-depth parsing and analysis of their logs, where the
parsing must correlate the log lines per-connection or per-queueid rather
than processing lines independently.  One of the barriers to this type of
processing is the unstructured nature of Postfix logs, where each log line
was added on an ad hoc basis as a requirement was discovered or new
functionality was added.\footnote{A history of all changes made to Postfix
is distributed with the source code, available from
\url{http://www.postfix.org/}} Further complication arises because the set
of rejection messages is not fixed: new messages can be added by the
administrator with custom checks; every \RBL{} returns a different
explanatory message; policy servers may log different messages depending on
the characteristics of the connection; there are many ways in which the log
lines may differ between servers, even within the same organisation when
servers are configured differently, or are running different version of
Postfix.  This paper documents the difficult process of parsing Postfix
logs, presenting a program which parses logs and places the resulting data
into a database.  The gathered data can then be used to optimise current
anti-spam measures, provide a baseline to test new anti-spam measures
against, or to produce statistics showing how effective those measures are.
There are numerous other uses for such data: improving server performance
by identifying troublesome destinations and reconfiguring appropriately or
dedicating transports to those destinations; identifying regular high
volume uses (e.g.\ customer newsletters) and restricting those uses to
off-peak times; detecting virus outbreaks which propagate via email; as a
base for billing customers on a shared server.  Preserving the raw data
enables users to develop a multitude of uses far beyond those conceived of
by the author.

@@

Section~\ref{rules} discusses the parsing rules in detail, describing the
table in the database used to hold the rules and explaining their
structure, referring to an example rule and sample data it matches
successfully against, followed by a discussion of rule efficiency concerns.

@@

Appendix~\ref{other-parsers} discusses other parsers and why they were
deemed unsuitable for the task.

@@

Appendix~\ref{graphs} contains graphs illustrating the topics previously
discussed in section~\ref{rule efficiency}, rule efficiency, and explains
the irregularities observed in the graphs.

Appendix~\ref{Glossary} provides a glossary of terms used in the paper.

@@

This section provides background information helpful in understanding the
remainder of the paper.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an \API\@; a brief introduction to
\ifacronymfirstuse{SMTP}{the}{} \SMTP\@; and a longer introduction to
Postfix, concentrating on the topics most relevant to this paper, namely
Postfix restrictions and policy servers.  The assumptions made in designing
and implementing the parser are explained, as are the conventions used in
this paper.  Other projects which attempt to parse Postfix logs are
summarised (full details are available in
section~\refwithpage{other-parsers}), plus a review of previously published
research in this area.

@@

These database queries yield summary statistics about the efficiency of
spam avoidance techniques that is far less feasible to assess directly from
log files without prior pre-processing into a database along the lines
proposed, implemented and tested herein.

@@

\subsection{SMTP background}
\label{SMTP background}

The \SMTPlong, originally defined in \RFC{} 821~\cite{RFC821} and updated
in \RFC{} 2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity many virus and/or spam sending programs fail to implement it
properly, so requiring strict adherence to the protocol specification is
beneficial in protecting against spam and viruses.\footnote{This violates
the principle of \textit{Be liberal in what you accept, and conservative in
what you send\/} from \RFC{} 760~\cite{rfc760}, but unfortunately that
principle was written in a friendlier time.}  A typical SMTP conversation
resembles the following (the lines starting with a three digit number are
sent by the server, all other lines are sent by the client):

@@

            \item Modify \texttt{smtpd\_data\_restrictions} in
                \texttt{/etc/postfix/main.cf} to contain\newline
                \hspace*{2em}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}

@@

        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
        place to log the HELO hostname, there won't be a queueid associated
        with the mail for the first recipient, so that log line cannot be
        associated with the correct mail.  There is guaranteed to be
        queueid when the DATA command has been reached, and thus it will be
        logged by any restrictions taking effect in
        \texttt{smtpd\_data\_restrictions}.

@@

% The appropriate command has been redefined so this will produce a section
% and label called glossary.
\printglossary{}
\printacronym{}
